---
title: "FDA Smoothing and Registration Notebook"
output: html_notebook
---

# Smoothing using Basis Decompositions and smoothness penalties

We illustrate the entire discussed procedure using log– precipitation data. This data was collected across several stations in Canada and is commonly used in the FDA literature to illustrate new methodologies.

```{r}
library(fda)
data(CanadianWeather)
attach(CanadianWeather)

y.precip=dailyAv[,,2]
l = which(place=="Vancouver") 
t.day = 1:365  
y=y.precip[,l]

#let's plot the data
plot(t.day, y, type="n",lwd=4, col="black",
     xlab="day", ylab="precipitation", 
     main="Vancouver Precipitation", cex=1)
points(t.day, y, pch=1, cex=.5, col="blue", lwd=1)

```

Let's introduce some smoothing using a spline basis, by specifying the number of basis elements:
```{r}
#Create a  spline basis over the interval (1,365) of order 3, with the specific number of basis
nbasis <- 10
ybasis  <- create.bspline.basis(rangeval = c(1,365), nbasis = nbasis, norder=4)

#do the smoothng
dayprecfd <- smooth.basis(t.day, y, ybasis)$fd 

main.label = paste("Vancouver (nbasis =", nbasis, ")", sep="")
plot(t.day, y, type="n", ylim=range(y), 
     ylab="Precipitation", xlab="day", main=main.label)
points(t.day, y, pch=1, cex=.5, col="blue", lwd=1)
lines(dayprecfd,col="red",lwd=4)
```

Now let's do it by introducing a penalty function.  We shall use int2Lfd(m) to introduce an m-derivative penalty term.  fdPar() : defines functional parameters; in this case the 2nd order derivative penalty term and the smoothing parameter.

```{r}
lambda <- 10^4

tD2fdPar = fdPar(ybasis, Lfdobj=int2Lfd(2), lambda=lambda)

# smooth.basis() : smoothes the data using the roughness penalty and smoothing parameter specified in 'tD2fdPar' 
dayprecfd = smooth.basis(t.day,y,tD2fdPar)$fd

main.label = paste("Vancouver (lambda =", round(lambda,2), ")", sep="")
plot(t.day, y, type="n", ylim=range(y), 
     ylab="Precipitation", xlab="day", main=main.label)
points(t.day, y, pch=1, cex=.5, col="blue", lwd=1)
lines(dayprecfd,col="red",lwd=4)
```

How do we select the optimal roughness penalty lambda? We can use cross-validation, generalized cross-validation, restricted maximum liklihood (reML) estimation, etc. The following code uses GCV to select optimal parameter from 50 candidates.

```{r}
loglambda=seq(-5, 12, len=71)  
#print the range  of the lambda
range(exp(loglambda))

gcv = rep(0,71)

for(i in c(1:length(loglambda))){
  lambda=exp(loglambda[i])
  
  tD2fdPar = fdPar(ybasis,Lfdobj=int2Lfd(2),lambda=lambda)
  tyfd = smooth.basis(t.day,y,tD2fdPar)
  
  gcv[i] = tyfd$gcv
}

# PLOT GCV of FIT versus log lambda
plot(loglambda,gcv[1:71],type='l',cex.lab=1.5, lwd=4, 
     xlab='log lambda',ylab='GCV', main="GCV(log.lambda)")
```

Let's find the minimum:

```{r}
index.logl.opt = which(gcv==min(gcv))
lambda.opt = exp(loglambda[index.logl.opt])
tD2fdPar = fdPar(ybasis,Lfdobj=int2Lfd(2),lambda=lambda.opt)
tyfd = smooth.basis(t.day,y,tD2fdPar)

plot(t.day, y, type="n", ylab="Precipitation", xlab="day", ylim=range(y), 
     main=paste("optimal lambda = ", round(lambda.opt)))
points(t.day, y, pch=1, cex=.5, col="blue", lwd=1)
lines(tyfd$fd,col="red",lwd=4)

```

# Derivatives

Once we have recovered a smooth curve associated with the functional data, we can also analyse rate of change properties, for example, places where the derivative  crosses the x-axis, etc. These ar crucial in identifying significant changes in trend -- however one must be careful that the derivatives are not too heavily influenced by a poor choice of smoothing parameters.  Once you obtain the derivative curves -- then confidence intervals etc can be derived and one can test whether the trend in a given region is significant.
```{r}
#  Compute the acceleration functions
dayprec_dt <- deriv.fd(dayprecfd, 1)
plot(dayprec_dt, type="n", ylab="Rate of Change Precipitation", xlab="day")
```

# Kernel Smoothing

Let us now try and do some kernel smoothing instead. We can use the ksmooth function.  The critical parameter here is the parameter h which controls the bandwidth

```{r}
  h = 100
  fit = ksmooth(t.day, y, kernel = c("box"), bandwidth = h)
  plot(t.day, y, type="n", ylab="Precipitation", 
       xlab="day", ylim=range(y), 
       main="Kernel Smoothing")
  points(t.day, y, pch=1, cex=.5, col="blue", lwd=1)
  lines(t.day, fit$y, col="red", lwd=2)
```

# Curve Registration

Let us experiment a bit with the continuous curve registration features offered in the fda package.  We shall re-produce the growth acceleration example in the notes, using the Berkely height dataset. 

```{r}
age=growth$age
heightBasis=create.bspline.basis( c(1,18), 35, 6, age) 
heightPar=fdPar(heightBasis, 3, 10^(-0.5)) 
heightSmooth=smooth.basis(age, growth$hgtf, heightPar)
plot(heightSmooth, lwd=2, xlab="age", ylab="height (cm)")
```

Now compute the accleration
```{r}
accelUnreg = deriv.fd(heightSmooth$fd, 2) 
plot(accelUnreg[,1], lwd=2, xlab="Age", ylab="Acceleration",ylim=c(-4,2)) 
mean.accelUnreg=mean(accelUnreg) 
lines(mean.accelUnreg, lwd=4, col="black")
```

Let's use the register.fd package to perform continuous registration
```{r}
regList=register.fd(yfd=accelUnreg) 
accelReg=regList$regfd
plot(accelReg, xlab="Age", ylab="Acceleration", ylim=c(-4,3))
```